### 2022-09-28 main/chapter_context.tex
wrap susy section

Gluons correspond to the eight generators of $SU\!(3)_C$, whose subscript $_C$
indicates that they operate on colour-charge triplets of quarks.

Weak and electromagnetic forces are unified in electroweak interactions with
four bosons $W^1$, $W^2$, $W^3$, and $B$ from the
$SU\!(2)_L$ and $U\!(1)_Y$ symmetries respectively.
These would be massless if the symmetry were perfectly respected, but it is
`spontaneously' broken with the Higgs mechanism~\cite{
higgs1964broken,
englert1964broken
}%
\footnote{%
Historical accuracy prefers the name `Brout–Englert–Higgs mechanism', but
Peter~Higgs has reportedly suggested
``ABEGHHKtH mechanism''~\cite{close2011infinity}
to fully credit other associated work.%
}
by adding four scalar fields (as a complex doublet) with a potential that moves
the ground state of to a gauge-\emph{asymmetric} configuration.
Around this new ground state, linear combinations of the electroweak fields
acquire masses while ``eating up''~\cite{rubakov1999classical} three of the
four scalar fields to yield the three weak bosons and the photon:
$W^1$ and $W^2$ mix into $W^+$ and $W^-$, while $W^3$ and $B$ mix into $Z$ and
$\gamma$.
The fourth scalar is the Higgs boson $h$~\cite{
glashow1959renorm,
weinberg1967model,
salam1959weak,
rubakov1999classical,
cottingham2007greenwood
}.

@@

Other mysteries of cosmology include:
the small positive cosmological constant,
the inflation of the early universe,
and baryogenesis~\cite{
wells2020discovery,
riess1998observational
}.
Answers to these mysteries all lie somewhere between uncertain and unknown,
but may not be entirely out of reach;
for example, the Higgs field could plausibly have driven
inflation~\cite{bezrukov2008higgs},
and the Standard Model does slightly violate $C$ and $CP$, both of which
are required for baryogenesis~\cite{sakharov1991re}.

@@

Certain Standard Model structures are quite general and parametrized such
that parts can be tuned to describe nature.
For example, the strong interaction includes a parameter $\theta_\mathrm{CP}$
whose non-zero values would induce $\mathrm{CP}$ violation in the strong
sector.
Observations show that it cannot be far from zero, and one promising feature of
axion-like models is their implication that $\theta_\mathrm{CP} = 0$~\cite{
kim2008axions,
thomson2013modern,
martin2017particle
}.

@@

% fermion boson multiplets
The MSSM matches each standard fermion to a pair of scalar
`sfermions' (left and right)
and each vector gauge field to a `gaugino' fermion
($\tilde g$, $\tilde W^1$, $\tilde W^2$, $\tilde W^3$, and $\tilde B$).
The Higgs field does not translate so directly, but is replaced with two sets
of complex scalar fields
($H^+_u$, $H^0_u$ and $H^0_d$, $H^\pm_d$) with
gaugino superpartners
($\tilde H^+_u$, $\tilde H^0_u$ and $\tilde H^0_d$, $\tilde H^-_d$).
(All sparticles are marked with tildes $\tilde{~}$.)
These fields mix into mass eigenstates which current data do not entirely
determine:
the gaugino mixtures are the charged charginos $C^\pm_i$ and
neutral neutralinos $N_i$
(also labelled $\chargino_i$,
$\neutralino_i$~\cite{atlas2022searches}),
and each squark and slepton has two states that mix the left and right fields.
From the four complex scalars, three degrees of freedom are eaten by the
weak bosons to leave the reborn Standard Model Higgs boson $h$
and four new scalars~\cite{
martin2016primer
}.

% made using:
% https://www.tablesgenerator.com/latex_tables
\begin{table}[tp]
\centering
\begin{tabular}{rrc}
\multirow{5}{*}{bosons}   & Higgs                         & \makebox[7em]{\hfill ($h$) \hfill $H^0$ \hfill $A^0$ \hfill $H^\pm\!\!$ \hfill}             \\[1ex]
                          & \multirow{2}{*}{squarks}      & \makebox[5em]{\hfill $\tilde u$ \hfill $\tilde c$ \hfill $\tilde t$ \hfill}                  \\
                          &                               & \makebox[5em]{\hfill $\tilde d$ \hfill $\tilde s$ \hfill $\tilde b$ \hfill}                  \\[1ex]
                          & \multirow{2}{*}{sleptions}    & \makebox[5em]{\hfill $\tilde e$ \hfill $\tilde \mu$ \hfill $\tilde \tau$ \hfill}             \\
                          &                               & \makebox[5em]{\hfill $\tilde \nu_e$ \hfill $\tilde \nu_\mu$ \hfill $\tilde \nu_\tau$ \hfill} \\[3ex]
\multirow{3}{*}{fermions} & gluinos                       & \makebox[5em]{\hfill $\tilde g$ \hfill}                                        \\[1ex]
                          & \multirow{2}{*}{gauginos}     & \makebox[5em]{\hfill $\tilde C^\pm_1\!\!$ \hfill $\tilde C^\pm_2\!\!$ \hfill} \\
                          &                               & \makebox[7em]{\hfill $\tilde N_1\!$ \hfill $\tilde N_2\!$ \hfill$\tilde N_3\!$ \hfill$\tilde N_4\!$ \hfill} \\
\end{tabular}
\caption[%
Particles and sparticles of the Minimal Supersymmetric Standard Model%
]{%
Particles and sparticles of the Minimal Supersymmetric Standard Model
in their mass eigenstates~\cite{martin2016primer}, in addition to the
Standard Model particles of Table~\ref{tab:context_particles_sm}.
The standard Higgs boson ($h$) is repeated here with its new pals.
Sparticles are marked with tildes $\tilde{~}$.
All listed bosons are spin-$0$ and all listed fermions are spin-$1/2$.
}
\label{tab:context_particles_mssm}
\end{table}

@@

% current observations reaching up to $\eV[T]$ scale
% hierarchy problem

Broken supersymmetry does not perfectly resolve the hierarchy problem, but
could still reduce its fine-tuning if the superpartner masses are below the
$\lesssim 1\,\eV[T]$ scale~\cite{martin2016primer}.

the MSSM has the success that it is consistent with the observed
$m(h) \lesssim 135\,\eV[G]$~\cite{
quiros1997constraints,
wells2020discovery
}

None of these hypothetical particles is yet observed, but it remains possible
that they could be invoked to describe future experimental data.

>>>>

In this model, gluons correspond to the eight generators of $SU\!(3)_C$, whose
subscript $_C$ indicates that they operate on colour-charge triplets of quarks.
Weak and electromagnetic forces are also unified into electroweak interactions
with four bosons $W^1$, $W^2$, $W^3$, and $B$ from the
$SU\!(2)_L$ and $U\!(1)$ symmetries respectively.
These would be massless if the symmetry were perfectly respected, but it is
`spontaneously' broken with the Higgs mechanism~\cite{
higgs1964broken,
englert1964broken
}%
\footnote{%
Historical accuracy prefers the name `Brout–Englert–Higgs mechanism', but
Peter~Higgs has reportedly suggested
``ABEGHHKtH mechanism''~\cite{close2011infinity}
to fully credit other associated work.%
}
by adding four scalar fields (as a complex doublet) with a potential that moves
the ground state of to a gauge-\emph{asymmetric} configuration.
Around this new ground state, linear combinations of the electroweak fields
acquire masses while ``eating up''~\cite{rubakov1999classical} three of the
four scalar fields to yield the three weak bosons and the photon:
$W^1$ and $W^2$ mix into $W^+$ and $W^-$, while $W^3$ and $B$ mix into $Z$ and
$\gamma$.
The fourth scalar is the Higgs boson $h$~\cite{
glashow1959renorm,
weinberg1967model,
salam1959weak,
rubakov1999classical,
cottingham2007greenwood
}.

@@

Other mysteries of cosmology include:
the small positive cosmological constant,
the inflation of the early universe,
and baryogenesis~\cite{
wells2020discovery,
riess1998observational
}.
Solutions to these mysteries all lie somewhere between uncertain and unknown,
but may not be entirely out of reach;
for example, the Higgs field could plausibly have driven
inflation~\cite{bezrukov2008higgs},
and the Standard Model does slightly violate $C$ and $CP$, both of which
are required for baryogenesis~\cite{sakharov1991re}.

@@

Certain Standard Model structures are quite general, and are parametrized such
that parts can be smoothly disabled if they do not describe nature.
For example, the strong interaction includes a parameter $\theta_\mathrm{CP}$
whose non-zero values would induce $\mathrm{CP}$ violation in the strong
sector.
Observations find that $\theta_\mathrm{CP}$ cannot be far from zero, and one
attractive feature of axion-like models is their requirement that
$\theta_\mathrm{CP} = 0$~\cite{
kim2008axions,
thomson2013modern,
martin2017particle
}.

@@

% fermion boson multiplets
The MSSM matches each standard fermion to a pair of scalar
`sfermions' (left and right)
and each vector gauge field to a `gaugino' fermion
($\tilde g$, $\tilde W^1$, $\tilde W^2$, $\tilde W^3$, and $\tilde B$).
The Higgs field does not translate so directly, but is replaced with two sets
of complex scalar fields
($H^+_u$, $H^0_u$ and $H^0_d$, $H^\pm_d$) with
gaugino superpartners
($\tilde H^+_u$, $\tilde H^0_u$ and $\tilde H^0_d$, $\tilde H^-_d$).
(All sparticles are marked with tildes $\tilde{~}$.)
These fields mix into mass eigenstates which current data do not entirely
determine:
the gaugino mixtures are the charged charginos $C^\pm_i$ and
neutral neutralinos $N_i$
(also labelled $\chargino_i$ and $\neutralino_i$~\cite{atlas2022searches}),
and each squark and slepton has two states that mix the left and right fields.
From the four complex scalars, three degrees of freedom are eaten by the
weak bosons to leave the reborn Standard Model Higgs boson $h$
and four new scalars~\cite{
martin2016primer
}.

% made using:
% https://www.tablesgenerator.com/latex_tables
\begin{table}[tp]
\centering
\begin{tabular}{rrc}
\multirow{5}{*}{bosons}   & Higgs                         & \makebox[7em]{\hfill ($h$) \hfill $H^0$ \hfill $A^0$ \hfill $H^\pm\!\!$ \hfill}             \\[1ex]
                          & \multirow{2}{*}{squarks}      & \makebox[5em]{\hfill $\tilde u$ \hfill $\tilde c$ \hfill $\tilde t$ \hfill}                  \\
                          &                               & \makebox[5em]{\hfill $\tilde d$ \hfill $\tilde s$ \hfill $\tilde b$ \hfill}                  \\[1ex]
                          & \multirow{2}{*}{sleptions}    & \makebox[5em]{\hfill $\tilde e$ \hfill $\tilde \mu$ \hfill $\tilde \tau$ \hfill}             \\
                          &                               & \makebox[5em]{\hfill $\tilde \nu_e$ \hfill $\tilde \nu_\mu$ \hfill $\tilde \nu_\tau$ \hfill} \\[3ex]
\multirow{3}{*}{fermions} & gluinos                       & \makebox[5em]{\hfill $\tilde g$ \hfill}                                        \\[1ex]
                          & \multirow{2}{*}{gauginos}     & \makebox[5em]{\hfill $\tilde C^\pm_1\!\!$ \hfill $\tilde C^\pm_2\!\!$ \hfill} \\
                          &                               & \makebox[7em]{\hfill $\tilde N_1\!$ \hfill $\tilde N_2\!$ \hfill$\tilde N_3\!$ \hfill$\tilde N_4\!$ \hfill} \\
\end{tabular}
\caption[%
Particles of the Minimal Supersymmetric Standard Model%
]{%
Particles of the Minimal Supersymmetric Standard Model
in their mass eigenstates~\cite{martin2016primer}, in addition to the
Standard Model particles of Table~\ref{tab:context_particles_sm}.
The standard Higgs boson ($h$) is repeated here with its new pals.
Sparticles are marked with tildes $\tilde{~}$.
All listed bosons are spin-$0$ and all listed fermions are spin-$1/2$.
}
\label{tab:context_particles_mssm}
\end{table}

@@

Supersymmetric modelling successfully predicted the lightest Higgs boson
mass to be below $\lesssim 130\,\eV[G]$ long before its observation at
$125\,\eV[G]$~\cite{
espinosa1992higgs,
espinosa1993upper,
quiros1997constraints,
wells2020discovery,
atlas2012higgs,
atlas2012combined,
cms2012higgs
}.
This prediction, however, assumed a mass scale below $1\,\eV[T]$,
and observations at the LHC are now limiting sparticle masses
into super-$\eV[T]$ territory (in specific models).
Such limits are also troublesome for the hierarchy problem, because large
superpartner masses introduce fine-tuning problems of their
own~\cite{martin2016primer}.
Despite its current lack of observational evidence, however,
supersymmetry remains a useful tool for building theoretically plausible
models, and might help us to understand future discoveries.
